Ricci curvature of the phylogenetic subtree-prune-regraft graph

### 2015-01-22 curvature.tex
lemma on min size of paired trees, rewrote curvature limit proof

\begin{lemma}
Let $T$ and $S$ be phylogenetic trees such that $\dspr{T,S}=1$, that is $T$ can be transformed into $S$ by moving a subtree $R$ with $k$ leaves.
The corresponding SPR operation moves $R$ from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$ excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degree of $T$ and $S$ differs by:
$$2(k(a-b) + i - j).$$

@@

Observe that the set of permissible moves changes after the movement of $R$ in four different ways: 1) Subtrees on the path from $U$ to $L$ may now be moved within $R$ and its newly introduced parent, 2) Subtrees on the path from $V$ to $L$ may no longer be moved within $R$ and its parent, 3) $R$'s parent subtree may now be moved within $U$, and 4) $R$'s parent subtree may no longer be moved within $V$.
Note that no additional moves are introduced are blocked by the SPR operation.

@@

@@

\begin{proposition}
Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence but with $R$ cut off and attached at a different location for each $T_n$.
Then $\lim_{n \rightarrow \infty} \kappa(T_n, S_n) = 0$ on the SPR graph.
\end{proposition}
\begin{proof}
First we show that $\kappa$ is bounded above by $(q_n+3(2k-1)(O(n)))/q_n$, where $q_n$ is the number of neighbors of $T_n$ (OK, this gets messy when $T_n$ and $S_n$ have different numbers of neighbors).
Let $T'_n$ be any SPR move away from $T_n$ where the SPR move does not cut $R$ into two and doesn't exclusively move $R$.
In the mass transport, we pair $T'_n$ with $S'_n$, which is the same move applied to $S_n$ (need to show this makes sense), such that $\dspr(T'_n, S'_n) = 1$.
This works for every tree that doesn't have $R$ cut in two.
The number of trees that \emph{do} have $R$ cut in two or moved by its root is $(2k-1)(O(n))$, and their distance is at most 3.

A lower bound is also available because that we can't do better than distance 1 for all $(T'_n, S'_n)$ pairs.
If we completely ignore the $(2k-1)(O(n))$ trees that cut $R$ in two, then we get a lower bound of $(q_n-(2k-1)(O(n)))/q_n$.

The proposition follows because $q_n$ is quadratic in $n$.
\end{proof}

>>>>

\begin{lemma}
\label{lem:degree_change}
Let $T$ and $S$ be phylogenetic trees such that $\dspr{T,S}=1$, that is $T$ can be transformed into $S$ by moving a subtree $R$ with $k$ leaves.
The corresponding SPR operation moves $R$ from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$ excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degree of $T$ and $S$ differs by:
$$2(k(a-b) + i - j).$$

@@

Observe that the set of permissible moves changes after the movement of $R$ in four different ways: 1) Subtrees on the path from $U$ to $L$ may now be moved within $R$ and its newly introduced parent, 2) Subtrees on the path from $V$ to $L$ may no longer be moved within $R$ and its parent, 3) $R$'s parent subtree may now be moved within $U$, and 4) $R$'s parent subtree may no longer be moved within $V$.
Note that no additional moves are introduced or blocked by the SPR operation.

@@

\begin{corollary}
\label{cor:paired_neighbors}
Let the degrees of $T$ and $S$ be $q$ and $p$ respectively.
Then at least $o = q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that:
\begin{enumerate}
\item $\dspr{T',S'} = 1$ for each $(T',S')$ pair, and
\item each neighbor of $T$ and each neighbor of $S$ is paired to exactly one tree.
\end{enumerate}
\end{corollary}
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target nodes.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
\end{proof}

@@

\begin{proposition}
Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence but with $R$ cut off and attached at a different location for each $T_n$.
Then $\lim_{n \rightarrow \infty} \kappa(T_n, S_n) = 0$ on the SPR graph.
\end{proposition}
\begin{proof}
First we show that $\kappa$ is bounded above by $(q_n+3(2k-1)(\OhOf{n}))/q_n$, where $q_n$ is the number of neighbors of $T_n$ and $S_n$.
Let $T'_n$ be any of the $o$ SPR moves from a pairing following Corollary~\ref{cor:paired_neighbors}.
In the mass transport, we pair $T'_n$ with $S'_n$, which is the same move applied to $S_n$ such that $\dspr{T'_n, S'_n} = 1$.
The number of trees unmatched by this pairing is $\OhOf{kn}$, and their distance is at most 3.
Thus, $\kappa$ is bounded above by $(q_n + 2\OhOf{kn}) / q_n$.

A lower bound is also available because we can't do better than distance 1 for all trees unless they move $R$.
There are only $\OhOf{n}$ pairs with trees that move $R$.
By ignoring these trees we get a lower bound of $(q_n -(\OhOf{n}))/q_n$.

The proposition follows because $q_n$ is quadratic in $n$.

Now, we prove the proposition when the number of neighbors differ.
Let $q_n$ and $p_n$ be the number of neighbors of $T_n$ and $S_n$, respectively.
Assume without loss of generality that $q_n < p_n$.
By Lemma~\ref{lem:degree_change}, $p_n - q_n = 2(k(a-b) + i - j)$, where $a,b,i,j < n$.
Thus, $p_n - q_n = \OhOf{kn}$.
We again pair $T'_n$ with $S'_n$ such that $\dspr{T'_n, S'_n}=1$ but, as $p_n > q_n$ we can only account for at most $q_n / p_n$ of the mass directly and may have to move the $(p_n - q_n) / p_n$ remainder to trees at distance at most 3.
Thus, $\kappa$ is bounded above by $(q_n + 3\OhOf{kn}) / p_n = (p_n + 2\OhOf{kn}) / p_n$.
We again bound $\kappa$ from below with $(q_n - \OhOf{n}) / q_n$ by ignoring the mass in trees that may overlap.
The proposition again follows because $q_n$ is quadratic in $n$.

\end{proof}
